---
title: "Life Expectancy Across Continents"
author: "Christa Ingabire"
date: 2026-09-21
---

## Introduction

For this post, I wanted to look at how life expectancy has changed across
different parts of the world. I used the Gapminder dataset, which contains
country-level data on life expectancy, population, and GDP per capita over
several decades.

I was mainly interested in two things: how continents compare in the latest
year available, and whether those differences look the same when viewed over
time.

Data source: [Gapminder](https://www.gapminder.org/). Free material from Gapminder, CC BY 4.0.

## Exploring the data

I started by loading the packages and looking at the first few rows of the
dataset.

```{r}
#| message: false
#| warning: false

library(gapminder)
library(ggplot2)

head(gapminder)
```

Each row represents a country in a particular year. For this post, I am mostly
interested in `continent`, `year`, and `lifeExp`

## Comparing continents in 2007

First, I checked the latest year available in the dataset and created a smaller
dataset containing only observations from that year.

```{r}
latest_year <- max(gapminder$year)
latest_data <- subset(gapminder, year == latest_year)

latest_year
```

The latest year is 2007. Next, I calculated the average life expectancy for
each continent

```{r}
latest_summary <- aggregate(
  lifeExp ~ continent,
  data = latest_data,
  FUN = mean
)

latest_summary
```

There is already a noticeable difference between the continents. Oceania has
the highest average life expectancy in 2007, followed by Europe, while Africa
has the lowest.

That made me curious about whether the gap between continents had always looked
this way, so I looked at the full time period next.

## Looking over time

To compare the trends, I calculated average life expectancy for each continent
in each year.

```{r}
continent_year <- aggregate(
  lifeExp ~ continent + year,
  data = gapminder,
  FUN = mean
)

head(continent_year)
```

Now each row represents the average for one continent in one year. I plotted
those averages to make the changes easier to compare.

```{r}
#| label: fig-life-expectancy
#| fig-cap: "Average life expectancy by continent from 1952 to 2007."

ggplot(
  continent_year,
  aes(x = year, y = lifeExp, color = continent)
) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  labs(
    title = "Life expectancy has risen across every continent",
    subtitle = "Average life expectancy, 1952–2007",
    x = NULL,
    y = "Life expectancy (years)",
    color = "Continent"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 11),
    legend.position = "bottom"
  )
```

## What I noticed

The graph gives a much better picture than the 2007 averages alone. Life
expectancy rises across every continent, but the paths are noticeably
different.

Asia has one of the biggest changes, starting below 50 years and reaching
above 70 by 2007. Europe and Oceania stay relatively high throughout the
dataset, while the Americas also show a fairly steady increase.

Africa stands out because its trend is less consistent. Life expectancy rises
for much of the period, but the line becomes much flatter around the 1990s and
early 2000s before increasing again toward 2007.

Looking at the full timeline was more useful than only comparing the final
year. The averages generally moved upward, but the differences between
continents were still visible at the end of the dataset.