---
title: "Life Expectancy Across Continents"
author: "Christa Ingabire"
date: 2026-09-21
---


## Introduction

Life expectancy is one way of summarizing differences in population health
across countries and over time. In this analysis, I use the Gapminder dataset
to explore how average life expectancy differs across continents and how those
differences changed between 1952 and 2007.

The analysis focuses on two questions:

1. How does average life expectancy compare across continents in the most
   recent year available?
2. How has average life expectancy changed across continents over time?

Data source: [Gapminder](https://www.gapminder.org/)

## Exploring the data

The Gapminder dataset contains country-level observations for population,
life expectancy, and GDP per capita across multiple years. I first load the
packages and inspect the first few observations

```{r}
#| message: false
#| warning: false

library(gapminder)
library(ggplot2)

head(gapminder)
```

Each row represents a country in a particular year. For this analysis, the
most important variables are continent, year, and lifeExp

## Life expectancy in the most recent year

Before comparing continents, I identify the most recent year represented in
the dataset

```{r}
latest_year <- max(gapminder$year)

latest_year
```

The most recent year in the dataset is 2007. I then restrict the data to that
year and calculate the mean life expectancy for each continent

```{r}
latest_summary <- gapminder |>
  subset(year == latest_year) |>
  aggregate(lifeExp ~ continent, data = _, FUN = mean)

latest_summary
```

The results show substantial differences in average life expectancy across
continents in 2007. Oceania has the highest average life expectancy, followed
by Europe, while Africa has the lowest average among the continents represented
in the dataset

However, this comparison only shows one point in time. Looking across the full
time period gives a better picture of how these differences developed

## Life expectancy over time

To examine change over time, I calculate the mean life expectancy for every
continent in each year

```{r}
continent_year <- aggregate(
  lifeExp ~ continent + year,
  data = gapminder,
  FUN = mean
)

head(continent_year)
```

This produces one average for each continent-year combination, allowing the
changes in life expectancy to be compared across the full period

```{r}
#| label: fig-life-expectancy
#| fig-cap: "Average life expectancy by continent from 1952 to 2007."

ggplot(
  continent_year,
  aes(x = year, y = lifeExp, color = continent)
) +
  geom_line(linewidth = 1) +
  labs(
    x = "Year",
    y = "Average life expectancy",
    color = "Continent"
  ) +
  theme_minimal()
```

## Findings

Average life expectancy generally increases across all five continents over
the period shown. The size and pattern of that increase, however, differ
between continents

Oceania and Europe remain among the continents with the highest average life
expectancy throughout the period. Asia shows a substantial increase, moving
from an average below 50 years in the early part of the dataset to above 70
years by 2007. The Americas also show a steady upward trend

Africa follows a different pattern. Average life expectancy increases for
much of the period but changes relatively little during the 1990s and early
2000s before increasing again by 2007. It remains below the averages of the
other continents throughout the dataset

## Conclusion

The Gapminder data show that average life expectancy increased across all
continents between 1952 and 2007, but the differences between continents did
not disappear. The analysis also demonstrates why examining a trend over time
can reveal information that a single-year comparison cannot.

These results are descriptive. They show patterns in the Gapminder data but
do not establish why the differences between continents exist or what caused
life expectancy to change over time.