---
title: "Passing Data Between R and Python"
author: "Christa Ingabire"
date: 2026-09-22
---

## R and Python in the same document

This post explores how R and Python can work together using `reticulate`.
I create data in R, use it in Python, and then bring a Python result back
into R for visualization.

First, I create a small data frame in R.

```{r}
#| message: false
#| warning: false

library(reticulate)
use_python(Sys.which("python"), required = TRUE)

practice <- data.frame(
  language = c("Python", "R", "SQL"),
  hours = c(8, 5, 6)
)

practice
```

Next, Python reads the R object using `r.practice` and calculates the number
of minutes practiced.

```{python}
practice_from_r = r.practice
practice_from_r
```

Back in R, I ask Python to perform the calculation in its main session. The
result can then be accessed from R using `py$minutes`.

```{r}
py_run_string("
practice_from_r = r.practice
minutes = practice_from_r['hours'] * 60
")

practice$minutes <- py$minutes

practice
```

The original data was created in R, Python calculated the new `minutes`
values, and R retrieved the Python result.

Finally, I use those values to create a simple graph in R.

```{r}
#| label: fig-practice-time
#| fig-cap: "Practice time by programming language, with minutes calculated in Python."

barplot(
  practice$minutes,
  names.arg = practice$language,
  xlab = "Language",
  ylab = "Practice time (minutes)"
)
```

This demonstrates data moving between the two languages. Python accessed the
R object through `r.practice`, and R accessed a result created in Python
through `py$minutes`.